Optimizing Abstract Abstract Machines

### 2012-11-13 paper/paper.tex

%\newpage
\section{Baseline analyzer}
\label{sec:baseline}

@@

\subsection{Baseline evaluation}
\label{sec:baselineeval}

The final approximation we make to get to our baseline analysis is
store-allocating results of application sub-expressions. Notice that
the $\kfn{}$ continuation stores a value - this makes the space of
continuations quadratic rather than linear in 0-CFA. We instead
allocate an address for the value position when we create the
continuation. This address and the tail address are both determined by
the label of the application point, so the space becomes linear and
the overall complexity drops to cubic. Note that this is a critical
abstraction in languages with n-ary functions, since otherwise the
continuation space becomes ${\mathcal O}(n^n)$!

@@

We discuss a variety of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation. They
come in two flavors: decreasing the size of the represented state
space [SS], and decreasing intermediate data structures [DS].

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector in
   the 0-CFA case. (preallocation)}
 \item{[SS] coarsely abstract literal data (abstracting compound data)}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors. The first three remain in the purely functional
realm. The final three use mutation in small, localized places, so
they do not increase conceptual complexity drastically.

@@

We can eliminate the {\tt ev} states from the run-time interpretation
of a program by specializing the machine transition relation to the
program being analyzed. This eliminates interpretative overhead by
first compiling the program into ``byte code'' instructions.

@@

Each step in the fixpointer for the above techniques requires joining
several, commonly large, stores together. Even with immutable hashes
that utilize sharing, this is memory-intensive. Worse, this
contributes an obscene fraction of the runtime. If we instead track
the changes to the store made over all steps and afterward update the
store, we end up with much less allocation and computational overhead.

We represent changes as a list of pairs of address to set of storeable
values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
each step. Applying the changes is straightforward:

@@

\begin{align*}
bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr' \\
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}

We also lift $\machstep$ differently to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated.

@@

The store delta technique removes overhead of joining large stores,
but for determining if a state has been visited before (vis. $f(X)
\setminus S$ in ${\mathcal G}$), we still have to compare entire stores - an
expensive computation.  We modify a technique created in Shivers'
dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
approximation, that removes the need for joining whole stores by
manipulating a global store imperatively. The original technique loses
precision because the starting store for each state in a machine
configuration can change between stepping all states and introduce
more flows than originally possible. However, by delaying the store
changes until after all steps have completed, we can maintain our
precision. In a variety of experiments, we observed that the overhead
for storing store-deltas was outweighed by the precision keeping flows
in check. That is to say, the analysis was faster {\it and} more
precise.

We pair this technique with a mutable workset. That is, to avoid the
intermediate sets introduced by the $State \to \Set(State)$ signature
of $\machstep$, we change the fixpoint algorithm to use a global
workset, {\tt todo}, that $\machstep$ imperatively updates with the
states that it reaches.

In total, we have six globals:
\begin{itemize}
 \item{{\tt $\msto$}, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from state to the timestamp at which it was last seen}
 \item{{\tt $\msdiff$}, the store changes for the current step}
 \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}

To ensure termination, we guard adding $c$ to {\tt todo} by the following check:
\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ and increase
{\tt T} as long as there was at least one change in $\msdiff$. This
logic is only sound if we know that each $\ttuple{\maddr}{\mval{s}}$
in $\msdiff$ would change the value of $\maddr$ in the current
store. Thus, we also guard additions to $\msdiff$ so that only updates
that would change the store are permitted. Each time $\msdiff$ is
successfully extended, we set {\tt $\msdiff$?} to true. Before each
individual step, we set it to false.

@@

This technique by itself only works in 0-CFA, since there are many more
addresses that can be allocated in 1-CFA and higher. To support
general $k$-CFA in a vector representation, we use a growable vector
and a map from addresses to assigned indices. When assignable indices
run out, we grow the vector. The cost of this map is less than a hash
representation for the whole store since typically there are more
references to an address than allocations of it. This is more
space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
up-front and addressing by the contour converted to polynomial, since
many combinations of labels are not likely to occur. The all up-front
allocation again breaks down when we add compound data, which
introduce new kinds of labels (e.g. left and right components of a
tuple).

\subsection{Abstracting Compound Data}

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
value of which do not particularly matter for control-flow
analysis. If we interpret a list literal of 30 elements (that again
have 10 elements each) as 300 applications of {\tt cons}, and then
interpret a function that traverses the tree, we will end up
interpreting that function at least 300 times, when once or twice
would do.

We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k-CFA$
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.

>>>>

\section{Reduction semantics to baseline analyzer}
\label{sec:baseline}

The uniform $k$-CFA allocation strategy would make the fixpoint of the
reduction relation in figure \ref{fig:aam} a computable analysis, but
not an effective one. This is not the strategy that AAM, nor we,
recommend. We explain a succession of approximations to reach the
baseline analysis that we compare performance against. But first, we
need a fixpoint combinator to drive the succession of reduction
relations we define.

@@

\subsection{Store-allocated results}
\label{sec:baselineeval}

The final approximation we make to get to our baseline analysis is
store-allocating results of application sub-expressions. The semantics
of the previous section is where AAM leaves off. However, notice that
the $\kfn{}$ continuation stores a value - this makes the space of
continuations quadratic rather than linear in 0-CFA. A linear space
drops the complexity to cubic.

To achieve this, we instead allocate an address for the value position
when we create the continuation.  This address and the tail address are
both determined by the label of the application point, so the space
becomes linear and the overall complexity drops to cubic. Note that
this is a critical abstraction in languages with n-ary functions,
since otherwise the continuation space becomes ${\mathcal O}(n^n)$!

@@

We discuss a succession of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation. They
come in two flavors: decreasing the size of the represented state
space [SS], and decreasing the amount of intermediate data structures [DS].

\begin{itemize}
 \item{[SS] lazy non-determinism: quotient by sets of values}
 \item{[SS] abstract compilation: eliminate {\tt ev} states}
 \item{[DS] store deltas: operate on store changes rather than whole stores}
 \item{[DS] timestamp: represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting}
 \item{[DS] preallocation: represent the global store as a global, mutable vector}
 \item{[SS] uniform literal approximation: coarsely abstract literal data}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors.

@@

Laziness removes several represented states that would then have to be
stepped by the fixpoint combinator.  We can further eliminate the {\tt
  ev} states from the run-time interpretation of a program by
specializing the machine transition relation to the program being
analyzed. This precision-preserving transformation eliminates
interpretative overhead by first compiling the program into ``byte
code'' instructions.

@@

Each step in the fixpoint for the above techniques requires joining
several, commonly large, stores together. Each step is unlikely to
always modify all addresses of the store, so this is needlessly
memory- and time-intensive. We can instead accumulate store changes
and apply them after all steps have completed. This uses far fewer
join operations, leading to much less allocation and computational
overhead.

We represent changes as $Store\msdiff = List[Addr \times
  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
begins as empty for each step. Applying the changes is
straightforward:

@@

\begin{align*}
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}
\begin{align*}
bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr'
\end{align*}
Compilation changes to additionally take a $\msdiff$ component, so the
above rule's right hand side would instead be $k^{\mcntr'}(\menv',
\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.

We also lift $\machstep$ to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated:

@@

The store delta technique removes overhead of joining large stores,
but for determining if a state has been visited before (vis. $f(X)
\setminus S$ in ${\mathcal G}$), we still have to compare entire
stores - an expensive computation.  We modify a technique due to
Shivers~\cite{ianjohnson:Shivers:1991:CFA}, timestamp approximation,
that removes the need for joining whole stores by manipulating a
global store imperatively. The original technique loses precision
because states are stepped in sequence, modifying the store in place -
more values or continuations can be added before another state gets to
step. This interference means more possible flows (i.e. less
precision) and a nondeterministic analysis - the order the set of states is
traversed is not necessarily deterministic. However, by delaying the
store changes until after all steps have completed, we can maintain
our precision and determinism. In a variety of experiments, we
observed that the time and space overhead for storing store-deltas was outweighed by
the precision keeping flows in check. That is to say, the analysis was
faster {\it and} more precise.

We pair this technique with a mutable workset to avoid the
intermediate sets introduced by the $State \times Store \to \Set(State) \times Store\msdiff$ signature
of $\machstep$. This necessitates a change to the fixpoint algorithm to use a global
workset, {\tt todo}, that $\machstep$ imperatively updates with the
states that it would have otherwise accumulated in a set that would later be added to the workset.

In total, we have six globals:
\begin{itemize}
 \item{{\tt $\msto$}, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
 \item{{\tt $\msdiff$}, the store changes for the current step}
 \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}

The new fixpoint calculation is defined in
figure~\ref{fig:imperative}.  To ensure termination, we guard adding
$c$ to {\tt todo} by the following check:
\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ imperatively and increase
{\tt T} as long as there was at least one change in $\msdiff$. This
logic leads to termination if we know that each $\ttuple{\maddr}{\mval{s}}$
in $\msdiff$ would change the value of $\maddr$ in the current
store. Thus, we also guard additions to $\msdiff$ so that only updates
that would change the store are permitted. Each time $\msdiff$ is
successfully extended, we set {\tt $\msdiff$?} to true. Before each
individual step, we set it to false.

@@

This technique by itself only works in 0-CFA, since there are many more
addresses that can be allocated in 1-CFA and higher. To support
general $k$-CFA in a vector representation, we use a growable vector
and a map from addresses to assigned indices. When assignable indices
run out, we grow the vector. The cost of this map is less than a hash
representation for the whole store since typically there are more
references to an address than allocations of it. This is more
space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
up-front and addressing by the contour converted to polynomial, since
many combinations of labels are not likely to occur. The all up-front
allocation again breaks down when we add compound data, which
introduce new kinds of labels (e.g. left and right components of a
tuple at the contour of creation).

\subsection{Abstracting literal compound data}

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
values of which do not particularly matter for many analyses. If we
interpret a list literal of 30 elements (that again have 10 elements
each) as 300 applications of {\tt cons}, and then interpret a function
that traverses the tree, we will end up interpreting that function at
least 300 times, when once or twice would do.

We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k$-CFA
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.
